---
title: "Corona Watch RIVM Data"
author: "Joeri Jongbloets"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE
  
)

library("EpiEstim")
library("tidyverse")
```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load helper functions
source(file.path(lib.dir, "trim.R"))
source(file.path(lib.dir, "estimate_r.R"))

# load corona watch data
source(file.path(data.dir, "cw.R"))

df.cw <- load_cw()
df.cw.latest <- df.cw %>%
  filter(report_date == max(report_date))

si_mean <- 3.96
si_sd <- 4.75
```

**Last timepoint: `r df.cw.latest %>% pull(date) %>% max()`**

```{r}

f_new_cases_day <- function(.d, counts_from = new_counts) {
  counts_from <- enquo(counts_from)
  
  .d %>%
    ggplot(aes(x = date, y=!!counts_from, colour= metric )) +
    facet_wrap(~metric, scales="free_y") +
    geom_smooth(se=F, method = "loess", formula = y ~ x) +
    geom_point() +
    scale_y_log10() +
    labs(
      x = "Date", y = "new #"
    ) +
    theme(legend.position = "bottom")
}

f_cases_day <- function(.d, counts_from = new_counts) {
  counts_from <- enquo(counts_from)
  
  .d %>%
    group_by(metric) %>%
    arrange(date) %>%
    mutate(
      counts = cumsum(!!counts_from)
    ) %>%
    ggplot(aes(x = date, y=counts, colour= metric )) +
    facet_wrap(~metric, scales="free_y") +
    geom_smooth(se=F, method = "loess", formula = y ~ x) +
    geom_point() +
    scale_y_log10() +
    labs(
      x = "Date", y = "total #"
    ) +
    theme(legend.position = "bottom")
}

```

## Raw Data

### Cases per Day

```{r}
df.cw.latest %>%
  f_new_cases_day()
```

### Total Cases over Time

```{r}
df.cw.latest %>%
  f_cases_day()
```

### Death Rate

```{r}
f_death_rate <- function(.d) {
  .d %>%
    group_by(metric) %>%
    arrange(date) %>%
    mutate(
      counts = cumsum(new_counts)
    ) %>%
    select(-new_counts) %>%
    pivot_wider(
      names_from = metric,
      names_prefix = "total_",
      values_from = counts
    ) %>%
    mutate(
      death_rate_cases = total_deaths / total_cases,
      death_rate_hospital = total_deaths / total_hospital
    )
}
```

```{r}
df.cw.latest %>%
  f_death_rate() %>%
  ggplot(aes(x=date, y=death_rate_cases)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Date",
    y = "Deaths per Case"
  )
```

```{r}
df.cw.latest %>%
  f_death_rate() %>%
  ggplot(aes(x=date, y=death_rate_hospital)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Date",
    y = "Deaths per Hospital Patient"
  )
```

### Effective R

```{r}
f_plot_r_over_time <- function( d ) {
  
  d %>%
    ggplot(aes(x=date, y=r_mean )) +
    geom_point(aes(colour = metric)) +
    geom_line(aes(colour = metric)) +
    geom_ribbon(aes(ymin = r_ci_lower, ymax = r_ci_upper, fill = metric), alpha=0.1) +
    geom_hline(aes(yintercept = 1), colour="red") +
    expand_limits(y=0) +
    labs(
      x = "Date",
      y = "Estimated R"
    )
  
}
```

```{r}
df.r <- df.cw.latest %>%
  group_by(metric) %>%
  nest() %>%
  group_by(metric) %>%
  mutate(
    model = map(
      data, f_estimate_r, 
      method = "parametric_si",
      cases_from = new_counts, time_from = date, 
      mean_si = si_mean, std_si = si_sd
    ),
    r_estimate = map2(data, model, f_extract_r, time_from = date)
  ) %>%
  unnest(r_estimate)
```

```{r}
df.r %>%
  f_plot_r_over_time
```

## Count Error

We now that counts are updated retrospectively, due to delays in reporting. Based on previous data we can  estimate the % error given the "age" of the value.

### Estimate Error

```{r}
f_update_probability <- function( .d ) {
  # .d = all counts for a given day
  
  t.first <- .d %>%
    filter(report_date == min(report_date)) %>%
    pull(report_date)
  
  final <- .d %>%
    filter(report_date == max(report_date)) %>%
    pull(new_counts)
  
  .d %>%
    arrange(report_date) %>%
    mutate(
      age = as.numeric(difftime(report_date, t.first, units = "days")),
      final_counts = final,
      perc_error = 1 - (new_counts / final_counts),
      perc_error = ifelse(final_counts == 0, 0, perc_error)
    )
  
}
```

```{r}
df.cw.error <- df.cw %>%
  group_by(metric, date) %>%
  nest() %>%
  mutate(
    new_data = map(data, f_update_probability),
  ) %>%
  select(-data) %>%
  unnest(new_data) %>%
  group_by(metric)
```

```{r}
df.cw.error %>%
  group_by(metric, age) %>%
  ggplot(aes(x=age, y =perc_error)) +
  facet_wrap(~metric) +
  geom_line(aes(colour=age, group=date), alpha=0.3) +
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon", alpha=0.5) +
  stat_summary(fun.y = "mean", geom = "line", colour="red") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Days since first report",
    y = "% Error"
  ) +
  theme(legend.position = "none")
```

```{r}

df.cw.error.model <- df.cw.error %>%
  group_by(metric, age) %>%
  summarise(
    mean_error = mean(perc_error),
    sd_error = sd(perc_error)
  )

f_apply <- function(new_counts, .age, .metric, df.model) {
  
  df.result <- df.model %>%
    filter(metric == .metric, age == .age) 
  
  if (nrow(df.result) == 0) {
    df.result <- df.model %>%
      group_by(metric) %>%
      filter(metric == .metric, age == max(age))
  }
  
  df.result %>% ungroup() %>% select(-c(metric, age))
}

df.cw.latest.f <- df.cw.latest %>%
  mutate(
    age = as.numeric(difftime(max(date), date, units="days")),
    prediction = pmap(
      list(new_counts=new_counts, .age=age, .metric=metric), 
      f_apply,
      df.model = df.cw.error.model
    )
  ) %>%
  unnest(prediction) %>%
  mutate(
    # E = 1 - (new_counts / final_counts)
    # final_counts = new_counts / (-E + 1)
    update_new_counts = new_counts / (-mean_error + 1)
  )
```

### Cases per Day

How much did the data change?

```{r}
df.cw.latest.f %>%
  pivot_longer(
    cols = c(new_counts, update_new_counts),
    names_to = "source",
    values_to = "counts",
  ) %>%
  mutate(
    source = ifelse(source == "new_counts", "original", "updated")
  ) %>%
  ggplot(aes(x = date, y=counts, shape=source, colour= metric )) +
  facet_wrap(~metric, scales="free_y") +
  # geom_smooth(se=F, method = "loess", formula = y ~ x) +
  geom_point() +
  geom_line(aes(group = date)) +
  scale_y_log10() +
  labs(
    x = "Date", y = "new #"
  ) +
  theme(legend.position = "bottom")
```

Only new data

```{r}
df.cw.latest.f %>%
 f_new_cases_day(counts_from = update_new_counts)
```

### Total Cases over time

```{r}
df.cw.latest.f %>%
  f_cases_day(counts_from = update_new_counts)
```

### Effective R

```{r}
df.r.f <- df.cw.latest.f %>%
  group_by(metric) %>%
  nest() %>%
  group_by(metric) %>%
  mutate(
    model = map(
      data, f_estimate_r, 
      method = "parametric_si",
      cases_from = update_new_counts, time_from = date, 
      mean_si = si_mean, std_si = si_sd
    ),
    r_estimate = map2(data, model, f_extract_r, time_from = date)
  ) %>%
  unnest(r_estimate)
```

```{r}
df.r.f %>%
  f_plot_r_over_time
```