---
title: "ECDC COVID-19 Data"
author: "Joeri Jongbloets"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning=FALSE
  
)

library("EpiEstim")
library("patchwork")
library("tidyverse")
```

```{r load_data, include=FALSE}
# Load project settings
source(here::here("settings.R"))

# load helper functions
source(file.path(lib.dir, "trim.R"))
source(file.path(lib.dir, "estimate_r.R"))

# load data
source(file.path(data.dir, "ecdc.R"))

df.ecdc <- load_ecdc()

si_mean <- 5
si_sd <- 3.4

```

* Last **global** timepoint: `r df.ecdc %>% pull(date_reported) %>% max()`
* Last **dutch** timepoint: `r df.ecdc %>% filter(geo_id == "NL") %>% pull(date_reported) %>% max()`

```{r}
f_plot_increase <- function(d) {
  d %>%
    filter( value > 0 ) %>%
    ggplot(aes(x = date_reported, y= value, colour=metric )) +
    facet_wrap(~metric, scales="free_y") +
    geom_smooth(se=F, method = "loess", formula = y ~ x) +
    geom_point() +
    scale_y_log10() +
    labs(
      x = "Date", y = "new #"
    )
}

f_plot_cumulative <- function(d) {
  d %>%
    group_by(metric) %>%
    arrange(date_reported) %>%
    mutate(
      cum_value = cumsum(value)
    ) %>%
    filter( cum_value > 0 ) %>%
    ggplot(aes(x = date_reported, y= cum_value, colour= metric )) +
    facet_wrap(~metric, scales="free_y") +
    geom_smooth(se=F, method = "loess", formula = y ~ x) +
    geom_point() +
    scale_y_log10() +
    labs(
      x = "Date", y = "cumalative #"
    )
}

f_plot_r_over_time <- function( d ) {
  
  d %>%
    ggplot(aes(x=date_reported, y=r_mean)) +
    facet_wrap(~metric) +
    geom_point(aes(colour = geo_id)) +
    geom_line(aes(colour = geo_id)) +
    geom_ribbon(aes(ymin = r_ci_lower, ymax = r_ci_upper, fill = geo_id), alpha=0.1) +
    geom_hline(aes(yintercept = 1), colour="red") +
    expand_limits(y=0) +
    labs(
      x = "Date",
      y = "Estimated R"
    )
  
}

f_plot_r_over_incidence <- function( d ) {
  
  d %>%
    group_by(metric) %>%
    arrange(date_reported) %>%
    mutate(
      cum_value = cumsum(value)
    ) %>%
    filter( cum_value > 0 ) %>%
    ggplot(aes(x=(cum_value / pop_size) * 1000, y=r_mean, colour=geo_id)) +
    facet_wrap(~metric, scales="free_x") +
    geom_point() +
    geom_line() +
    # scale_x_log10() +
    expand_limits(y=0) +
    labs(
      x = "# cases per 1000",
      y = "Estimated R"
    )
}

```

```{r}
df.ecdc %>%
  group_by(country) %>%
  summarise(
    last_report_date = max(date_reported)
  ) %>%
  ggplot(aes(last_report_date)) +
  geom_histogram()
```

## Cases per Day

### Global Situation

```{r}
df.ecdc %>%
  summarise_ecdc %>%
  f_plot_increase()
```

```{r}
df.ecdc %>%
  summarise_ecdc %>%
  f_plot_cumulative()
```

### European Situation

```{r}
df.ecdc %>%
  filter(geo_id %in% european) %>%
  summarise_ecdc %>%
  f_plot_increase()
```

```{r}
df.ecdc %>%
  filter(geo_id %in% european) %>%
  summarise_ecdc() %>%
  f_plot_increase()
```

```{r}
df.ecdc %>%
  filter(geo_id %in% european) %>%
  summarise_ecdc %>%
  f_plot_cumulative()
```

```{r}
df.ecdc.europe <- df.ecdc %>%
  filter(geo_id %in% european) %>%
  group_by(metric, geo_id) %>%
  summarise(
    value = sum(value)
  ) %>%
  group_by(metric) %>%
  mutate(
    perc_value = value / sum(value),
  )

p1 <- df.ecdc.europe %>%
  filter(metric == "cases") %>%
  ggplot(aes(x=fct_reorder(geo_id, perc_value), y=perc_value, fill = geo_id)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Country",
    y = "% of total"
  )

p2 <- df.ecdc.europe %>%
  filter(metric == "deaths") %>%
  ggplot(aes(x=fct_reorder(geo_id, perc_value), y=perc_value, fill = geo_id)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Country",
    y = "% of total"
  )


p1 + p2 + plot_layout(guides = 'collect') & theme(legend.position = 'none')
```

### Italian Situation

```{r}
df.ecdc %>%
  filter(geo_id == "IT") %>%
  f_plot_increase()
```

```{r}
df.ecdc %>%
  filter(geo_id == "IT") %>%
  f_plot_cumulative()
```

### Dutch Situation

```{r}
df.ecdc %>%
  filter(geo_id == "NL") %>%
  f_plot_increase()
```

```{r}
df.ecdc %>%
  filter(geo_id == "NL") %>%
  f_plot_cumulative()
```

### Overlap

```{r}
df.params <- data.frame(
  geo_id = c("NL", "IT"),
  date_start = c( "2020-02-27", "2020-02-21" ),
  population = c(17424978 ,60549600)
) %>%
  mutate(
    date_start = lubridate::ymd(date_start)
  )
```

```{r}
df.ecdc %>%
  filter(geo_id %in% c("NL", "IT")) %>%
  left_join(
    df.params, by = c("geo_id")
  ) %>%
  mutate(
    t_day = as.numeric(difftime(date_reported, date_start, units = "days" )),
    value_norm = (value / population) * 17424978
  ) %>%
  filter( t_day > 0 ) %>%
  ggplot(aes(x = t_day, y= value_norm, colour= geo_id )) +
  facet_wrap(~metric, scales="free_y") +
  geom_smooth(se=F) +
  geom_point() +
  scale_y_log10() +
  labs(
    x = "Date", y = "Normalized new #"
  )
```

```{r}
df.ecdc %>%
  filter(geo_id %in% c("NL", "IT")) %>%
  left_join(
    df.params, by = c("geo_id")
  ) %>%
  group_by(geo_id, metric) %>%
  arrange(date_reported) %>%
  mutate(
    cum_value = cumsum(value)
  ) %>%
  ungroup() %>%
  mutate(
    t_day = as.numeric(difftime(date_reported, date_start, units = "days" )),
    cum_value_norm = (cum_value / population) * 17424978
  ) %>%
  filter( t_day > 0 ) %>%
  ggplot(aes(x = t_day, y= cum_value_norm, colour= geo_id )) +
  facet_wrap(~metric, scales="free_y") +
  geom_smooth(se=F) +
  geom_point() +
  scale_y_log10() +
  labs(
    x = "Date", y = "Normalized cumalative #"
  )
```

## Death Rate

```{r}
f_plot_death_rate <- function( .d ) {
  
  .d %>%
    pivot_wider(
      names_from = metric,
      names_prefix = "new_",
      values_from = value
    ) %>%
    arrange(date_reported) %>%
    mutate(
      cum_cases = cumsum(new_cases),
      cum_deaths = cumsum(new_deaths),
      death_rate = cum_deaths / cum_cases
    ) %>%
    filter(cum_cases > 0, cum_deaths > 0) %>%
    ggplot(aes(x=date_reported, y=death_rate)) +
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = scales::percent) +
    expand_limits(y=0) +
    labs(
      x = "Date",
      y = "Death Rate"
    )
  
}
```

### Global Situation

```{r}
df.ecdc %>%
  summarise_ecdc %>%
  f_plot_death_rate()
```

### European Situation

```{r}
df.ecdc %>%
  filter(geo_id %in% european) %>%
  summarise_ecdc %>%
  f_plot_death_rate()
```

### Italian Situation

```{r}
df.ecdc %>%
  filter(geo_id == "IT" ) %>%
  f_plot_death_rate()
```

### Dutch Situation

```{r}
df.ecdc %>%
  filter(geo_id == "NL") %>%
  f_plot_death_rate()
```

## Estimate R

```{r}
df.r <- df.ecdc %>%
  group_by(geo_id, metric) %>%
  nest() %>%
  bind_rows(
    df.ecdc %>%
      group_by(metric) %>%
      summarise_ecdc %>%
      mutate(geo_id = "Global") %>%
      group_by(metric, geo_id) %>%
      nest(),
    df.ecdc %>%
      filter(geo_id %in% european) %>%
      group_by(metric) %>%
      summarise_ecdc %>%
      mutate(geo_id = "EU") %>%
      group_by(metric, geo_id) %>%
      nest()
  ) %>%
  group_by(geo_id, metric) %>%
  mutate(
    model = map(data, f_estimate_r, cases_from = value, si_mean = si_mean, si_sd = si_sd),
    r_estimate = map2(data, model, f_extract_r, time_from = date_reported)
  ) %>%
  filter(!is.na(r_estimate))
```

### Global Situation

```{r}
df.r %>%
  filter(geo_id == "Global") %>%
  unnest(r_estimate) %>%
  f_plot_r_over_time()
```

### European Situation

```{r}
df.r %>%
  filter(geo_id == "EU") %>%
  unnest(r_estimate) %>%
  f_plot_r_over_time()
```

### Italian Situation

```{r}
df.r %>%
  filter(geo_id == "IT") %>% 
  unnest(r_estimate) %>%
  f_plot_r_over_time
```

### Dutch Situation

```{r}
df.r %>%
  filter(geo_id == "NL") %>% 
  unnest(r_estimate) %>%
  f_plot_r_over_time
```

### Overlap

```{r}
df.r %>%
  group_by(geo_id) %>%
  filter(geo_id %in% c("IT" , "NL", "DE")) %>% 
  unnest(r_estimate) %>%
  filter(t_day > -21) %>%
  f_plot_r_over_time
```

### R vs Incidence

```{r}
df.r %>%
  group_by(geo_id, metric) %>%
  filter(geo_id %in% c("IT" , "NL", "DE")) %>% 
  unnest(r_estimate) %>%
  filter(value > 0) %>% 
  f_plot_r_over_incidence() +
  coord_cartesian(ylim = c(0, 10))
```
